Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-09-03 paper.Rnw
made adjustments to figures to meet minimum size (text and lines) requiements for submission

# Specify global options
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.keep = 'high',
  fig.align = 'center',
  dev = c("cairo_pdf", "cairo_ps"),
  dev.args = list(fallback_resolution = 800),
  dpi = 800
)

# Load packages
library(Cairo)
library(caret)
library(cowplot)
library(dplyr)
library(forcats)
library(ggforce)
library(ggplot2)
library(ggpcp) #devtools::install_github("yaweige/ggpcp", build_vignettes = TRUE)
library(glmnet)
library(gower)
library(graphics)
library(gretchenalbrecht) #devtools::install_github("dicook/gretchenalbrecht")
library(latex2exp)
library(lime) #devtools::install_github("goodekat/lime")
library(limeaid) #devtools::install_github("goodekat/limeaid")
library(purrr)
library(randomForest)
library(stringr)
library(tidyr)

@@

<<concept-data>>=
# Simulate example data
set.seed(20190624)
lime_data <-
  data.frame(
    feature1 = sort(runif(250, 0, 1)),
    feature2 = sample(x = 1:250, size = 250, replace = FALSE)) %>%
  mutate(
    feature1_stnd = (feature1 - mean(feature1)) / sd(feature1),
    feature2_stnd = (feature2 - mean(feature2)) / sd(feature2),
    prediction = if_else(feature1 >= 0 & feature1 < 0.1,
                         (0.3 * feature1) + rnorm(n(), 0, 0.01),
                 if_else(feature1 >= 0.1 & feature1 < 0.3,
                         rbeta(n(), 1, 0.5),
                 if_else(feature1 >= 0.3 & feature1 < 0.5,
                         sin(pi* feature1) + rnorm(n(), 0, 0.5),
                 if_else(feature1 >= 0.5 & feature1 < 0.8,
                         -(sin(pi* feature1) + rnorm(n(), 0, 0.1)) + 1,
                 if_else(feature1 >= 0.8 & feature1 < 0.9,
                         0.5 + runif(n(), -0.5, 0.5),
                         0.5 + rnorm(n(), 0, 0.3))))))) %>%
  bind_rows(
    data.frame(feature1 = rep(1, 30) + rnorm(5, 0, 0.05),
           feature2 = rep(245, 30) + rnorm(5, 0, 1),
           prediction = rep(0, 30) + rnorm(5, 0, 0.05)) %>%
  mutate(feature1_stnd = (feature1 - mean(feature1)) / sd(feature1),
         feature2_stnd = (feature2 - mean(feature2)) / sd(feature2)))

@@

\begin{figure*}[!thp]
<<concept-plot-good, out.width = '6.5in', fig.width = 7.95, fig.height = 3.75, warning = FALSE>>=

# Plot of good explainer model
ggplot() +
  facet_grid(. ~ feature, switch = "x") +
  geom_point(data = lime_data_gathered,
             mapping = aes(x = feature_stnd_value,
                           y = prediction,
                           size = distance_value,
                           alpha = distance_value,
                           color = distance)) +
  geom_abline(data = coefs_data %>% filter(case == "good"),
              mapping = aes(intercept = int, slope = slope),
              size = 1) +
  scale_alpha(range = c(0.3, 1)) +
  scale_color_manual(values = c(NA, "grey30"), guide = "none") +
  geom_point(data = prediction_of_interest_gathered,
             mapping = aes(x = feature_value,
                           y = prediction,
                           fill = color),
             color = "black",
             size = 5,
             shape = 23,
             alpha = 0.75) +
  scale_fill_manual(values = "#FAAA72") +
  geom_text(
    data = data.frame(feature_stnd_value = 1.2,
                      prediction = 1.6,
                      feature = c("Feature 1", "Feature 2"),
                      slope = c(paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "good",
                                               feature == "Feature 1") %>%
                                        pull(slope) %>%
                                        round(3)),
                                paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "good",
                                               feature == "Feature 2") %>%
                                        pull(slope) %>%
                                        round(3)))),
    mapping = aes(x = feature_stnd_value,
                  y = prediction,
                  label = slope),
    family = "Times",
    size = 4) +
  labs(x = "",
       y = "Black-Box Prediction",
       title = "Conceptual Depiction of a Faithful Local Explainer Model",
       subtitle = paste("Gower Distance Metric Exponent:", good_gower_power),
       fill = "",
       alpha = "Weight",
       size = "Weight") +
  theme_linedraw(base_family = "Times", base_size = 8) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.placement = "outside",
        strip.background = element_rect(color = "white", fill = "white"),
        strip.text = element_text(color = 'black', size = 10)) +
  guides(fill = guide_legend(order = 1),
         size = guide_legend(order = 2),
         alpha = guide_legend(order = 2))

@
\caption{A conceptual depiction of a faithful LIME explainer model in the immediate neighborhood of a prediction of interest. The predictions from a hypothetical black-box model are plotted against the standardized values of the two hypothetical predictor variables. The diamond shaped points represent the location of a prediction of interest. The size and opacity of the circular points indicate the weight assigned based on the distance to the prediction of interest computed using the inverse of the Gower distance metric raised to the power of \Sexpr{good_gower_power}. The black lines represent  a weighted ridge regression model used as an explainer model that reasonably captures the relationship between the black-box predictions and the features in a local region around the prediction of interest. That is, the explainer is faithful to the complex model and produces a reasonable explanation that Feature 1 plays a more important role in the prediction of interest than Feature 2 since the magnitude of the slope associated with Feature 1 is larger.}
\label{fig:concept-plot-good}

@@

<<concept-plot-bad, out.width = '6.5in', fig.width = 7.95, fig.height = 3.75, warning = FALSE>>=

# Plot of good explainer model
ggplot() +
  facet_grid(. ~ feature, switch = "x") +
  geom_point(data = lime_data_gathered,
             mapping = aes(x = feature_stnd_value,
                           y = prediction,
                           size = distance_value,
                           alpha = distance_value,
                           color = distance)) +
  geom_abline(data = coefs_data %>% filter(case == "bad"),
              mapping = aes(intercept = int, slope = slope),
              size = 1) +
  scale_alpha(range = c(0.3, 1)) +
  scale_color_manual(values = c("grey30", NA), guide = "none") +
  geom_point(data = prediction_of_interest_gathered,
             mapping = aes(x = feature_value,
                           y = prediction,
                           fill = color),
             color = "black",
             size = 5,
             shape = 23,
             alpha = 0.75) +
  scale_fill_manual(values = "#FAAA72") +
  geom_text(
    data = data.frame(feature_stnd_value = 1.2,
                      prediction = 1.6,
                      feature = c("Feature 1", "Feature 2"),
                      slope = c(paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "bad",
                                               feature == "Feature 1") %>%
                                        pull(slope) %>%
                                        round(3)),
                                paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "bad",
                                               feature == "Feature 2") %>%
                                        pull(slope) %>%
                                        round(3)))),
    mapping = aes(x = feature_stnd_value,
                  y = prediction,
                  label = slope),
    family = "Times",
    size = 4) +
  labs(x = "",
       y = "Black-Box Prediction",
       title = "Conceptual Depiction of an Unfaithful Local Explainer Model",
       subtitle = paste("Gower Distance Metric Exponent:", bad_gower_power),
       fill = "",
       alpha = "Weight",
       size = "Weight") +
  theme_linedraw(base_family = "Times", base_size = 8) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.placement = "outside",
        strip.background = element_rect(color = "white", fill = "white"),
        strip.text = element_text(color = 'black', size = 10)) +
  guides(fill = guide_legend(order = 1),
         size = guide_legend(order = 2),
         alpha = guide_legend(order = 2))

@
\caption{A conceptual depiction of an unfaithful local explainer model in the immediate neighborhood of a prediction of interest. A ridge regression model is fit to the same hypothetical black-box model predictions and predictor variables as in \autoref{fig:concept-plot-good}, but the weights are computed using a Gower exponent of 1. The explainer model is unfaithful to the complex model in any immediate neighborhood of the prediction of interest.}
\label{fig:concept-plot-bad}
\end{figure*}

\autoref{fig:concept-plot-good} provides a visualization of this conceptual understanding of LIME. The two plots show the predictions from a hypothetical black-box model plotted against the two hypothetical model predictor variables. The diamond shaped points represent the location of the prediction of interest. A Gower distance metric \citep{gower:1971} is used to define locality such that the size of the points represent the inverse of the Gower distance raised to some value and indicate the proximity to the prediction of interest. In the example of \autoref{fig:concept-plot-good}, an exponent of \Sexpr{good_gower_power} is used to emphasize a very local region around the prediction of interest. A ridge regression model weighted by the proximity values is used as the explainer model with the black-box predictions as the response variable and standardized versions of two features in the data as predictor variables. Standardized features allow direct comparisons of the model coefficients. The explainer model is depicted by the black lines in the figure.

The plot on the left of \autoref{fig:concept-plot-good} shows the relationship between the black-box predictions and Feature 1. Here, the explainer model is plotted with Feature 2 fixed at the observed value of the prediction of interest. The explainer model captures the relationship in an  immediate neighborhood around the prediction of interest with a slope of \Sexpr{coefs_data %>% filter(case == "good", feature == "Feature 1") %>% pull(slope) %>% round(3)}. The plot on the right shows no global or local relationships between the black-box predictions and Feature 2. Here, the explainer model is plotted with Feature 1 set as the observed value of the prediction of interest, and it has an appropriately small slope of \Sexpr{coefs_data %>% filter(case == "good", feature == "Feature 2") %>% pull(slope) %>% round(3)}. The magnitude of the slope associated with Feature 1 is larger than the slope of Feature 2, which suggests that Feature 1 plays a more important role in the prediction made by the black-box model for the prediction of interest. This explanation agrees with a visual assessment of the relationships between the predictions and predictor variables.

@@

We consider an example where the choice of exponent for the weights is crucial to the quality of the explanation. The plots in Figure \ref{fig:concept-plot-bad} show the same data as \autoref{fig:concept-plot-good}, but the Gower distance metric exponent is decreased to \Sexpr{bad_gower_power} (the default exponent in the \emph{lime} R package). This causes the observations that are further away from the prediction of interest to be given larger weights than before. In \autoref{fig:concept-plot-good}, the explainer model captures the relationship between the black-box predictions in an immediate neighborhood of the prediction of interest. This cannot be said of the explainer model in \autoref{fig:concept-plot-bad}. In addition, the magnitude of the slope associated with Feature 2  (\Sexpr{coefs_data %>% filter(case == "bad", feature == "Feature 2") %>% pull(slope) %>% round(3) %>% abs()}) is larger than that of Feature 1 (\Sexpr{coefs_data %>% filter(case == "bad", feature == "Feature 1") %>% pull(slope) %>% round(3) %>% abs()}). Thus, this explainer model actually provides a misleading explanation that Feature 2 plays a more important role in the prediction of interest.

Several sources in the literature discuss the performance of LIME. One of the biggest difficulties with LIME is determining how to specify a local region \citep{laugel:2018} \citep{molnar:2019}. This is due to an unclear definition of a "local region" and how to apply LIME to achieve an appropriate local region as demonstrated by \autoref{fig:concept-plot-bad}. \citet{alvarezmelis:2018} raise a concern pertaining to the robustness of explanations from LIME and other model explainers: they find that even small changes in predictor variables can lead to very different LIME explanations. Additionally, \citet{ribeiro:2016} acknowledge that if a linear model is used as the explainer, LIME relies on a linear approximation of the explainer model to the complex model and state "if the underlying model is highly non-linear even in the locality of the prediction, there may not be a faithful explanation".

@@

\begin{figure*}[!thp]
\centering
<<sine-plot-data, out.width = '6.5in', fig.width = 8, fig.height = 3.1>>=

@@

# Plot the training data observed classes
sine_plot_obs <- ggplot(sine_data_train, aes(x = x1, y = x2, color = y)) +
  geom_point(alpha = 0.8) +
  geom_path(data = sinefun_data, aes(x = x, y = y),
            color = "black") +
  scale_colour_manual(values = c(sinecolor_blue, sinecolor_red)) +
  theme_bw(base_family = "Times", base_size = 8) +
  labs(x = TeX("$x_1$"),
       y = TeX("$x_2$"),
       color = TeX("y"),
       title = "Training Data")

# Plot the testing data rf predictions
sine_plot_pred <- ggplot() +
  geom_point(data = sine_poi %>%
               mutate(shape = "Prediction \nof Interest"),
             mapping = aes(x = x1,
                           y = x2,
                           #fill = rfprob_blue,
                           shape = shape),
             size = 5,
             alpha = 0.8,
             color = "black") +
  geom_point(data = sine_data_test %>% filter(y != rfpred),
             mapping = aes(x = x1, y = x2),
             shape = 1,
             size = 3,
             color = "black") +
  geom_point(data = sine_data_test,
             mapping =  aes(x = x1,
                            y = x2,
                            color = rfprob_blue,
                            fill = rfprob_blue),
             shape = 21,
             alpha = 0.8) +
  geom_path(data = sinefun_data, aes(x = x, y = y),
            color = "black") +
  scale_color_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5) +
  scale_fill_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5) +
  scale_shape_manual(values = 23) +
  theme_bw(base_family = "Times", base_size = 8) +
  labs(x = TeX("$x_1$"),
       y = TeX("$x_2$"),
       color = "Random \nForest \nProbability \nfor 'blue'",
       fill = "Random \nForest \nProbability \nfor 'blue'",
       title = "Testing Data",
       shape = "") +
  guides(shape = guide_legend(order = 1))

@@

@
\caption{Plots of $x_2$ versus $x_1$ from the training (left) and testing (right) sets of the \data \ introduced in Section \ref{diagnostics}. The true classification boundary is shown as the solid black line in both plots. The color of the training data  represents the value of the observed response variable ($y$). The color of the testing data  represents the random forest probability that an observation belongs to the category of blue. The \Sexpr{dim(sine_data_test %>% filter(y != rfpred))[1]} cases that are misclassified by the random forest are identified by black circles. The prediction of interest to explain is indicated by a diamond.}
\label{fig:sine-plot-data}
\end{figure*}

@@

# Extract the explanations from the default lime application
# for the prediction of interest only
sine_poi_lime_default <- sine_lime_default %>%
  filter(case == sine_poi %>% pull(case)) %>%
  mutate(case = c("Prediction of Interest", "Prediction of Interest"))
@

@@

\autoref{fig:sine-plot-data} shows scatterplots of $x_2$ versus $x_1$ from the training data (left) and the testing data (right). Both plots include the true classification boundary of the rotated sine function plotted as the solid black line. The training data are colored by the observed response variable ($y$), and the testing data are colored by random forest  prediction probabilities. The random forest  misclassifies \Sexpr{dim(sine_data_test %>% filter(y != rfpred))[1]} points, which are all located near  the classification boundary. These are identified by open circles in \autoref{fig:sine-plot-data}.

@@

For the presentation of the explanation scatterplot, we focus on the misclassified point with $(x_1, x_2, x_3)$ coordinates of (\Sexpr{round(sine_poi$x1, 2)}, \Sexpr{round(sine_poi$x2, 2)}, \Sexpr{round(sine_poi$x3, 2)}) indicated by a diamond in \autoref{fig:sine-plot-data}. Misclassified points are often of interest to explain since they may provide information about ways to improve the model. For the introduction of the other two  plots, we consider the LIME explanations for  all observations in the \data \ test data.

<<warning = FALSE>>=
# Obtain the simulated data associated with the poi
sine_poi_perms <- sine_poi_lime_default %>%
  slice(1) %>%
  select(perms_raw, perms_pred_complex, perms_numerified, weights) %>%
  mutate(perms_numerified =
           map(perms_numerified,
               .f = function(x) rename(x,
                                       "x1num" = "x1",
                                       "x2num" = "x2",
                                       "x3num" = "x3"))) %>%
  unnest(cols = c(perms_raw, perms_pred_complex, perms_numerified, weights)) %>%
  select(-red) %>%
  rename(rfpred = blue) %>%
  mutate(case = factor(1:n())) %>%
  select(case, everything())

# Determine the bin cuts for the default lime application
sine_lime_default_bin_cuts <- sine_lime_default %>%
  select(feature, feature_desc) %>%
  unique() %>%
  separate(feature_desc, c("other", "bin_cut"), sep = " <= ") %>%
  select(-other) %>%
  na.omit() %>%
  mutate(bin_cut = as.numeric(bin_cut)) %>%
  arrange(feature) %>%
  mutate(case = rep(1:3, 2)) %>%
  spread(feature, bin_cut)
@

@@

An explanation scatterplot corresponding to the LIME explanation depicted on the left in \autoref{fig:sine-exp-scatter} is shown on the right hand side of \autoref{fig:sine-exp-scatter}. By juxtaposing the random forest predictions and the explainer model boundaries, we are able to assess the faithfulness and localness of the explainer model.

>>>>

# Specify global options
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.keep = 'high',
  fig.align = 'center',
  dev = c("cairo_pdf", "cairo_ps"),
  dev.args = list(fallback_resolution = 800),
  dpi = 1200
)

# Load packages
library(Cairo)
library(caret)
library(cowplot)
library(dplyr)
library(forcats)
library(ggforce)
library(ggplot2)
library(ggpcp) #devtools::install_github("yaweige/ggpcp", build_vignettes = TRUE)
library(glmnet)
library(gower)
library(graphics)
library(gretchenalbrecht) #devtools::install_github("dicook/gretchenalbrecht")
library(latex2exp)
library(lime) #devtools::install_github("goodekat/lime")
library(limeaid) #devtools::install_github("goodekat/limeaid")
library(magick)
library(purrr)
library(randomForest)
library(stringr)
library(tidyr)

# Specify the (base) font size for all plots
fs = 7
@

@@

<<concept-data>>=

# Simulate example data
set.seed(20190624)
lime_data <-
  data.frame(
    feature1 = sort(runif(250, 0, 1)),
    feature2 = sample(x = 1:250, size = 250, replace = FALSE)) %>%
  mutate(
    feature1_stnd = (feature1 - mean(feature1)) / sd(feature1),
    feature2_stnd = (feature2 - mean(feature2)) / sd(feature2),
    prediction = if_else(feature1 >= 0 & feature1 < 0.1,
                         (0.3 * feature1) + rnorm(n(), 0, 0.01),
                 if_else(feature1 >= 0.1 & feature1 < 0.3,
                         rbeta(n(), 1, 0.5),
                 if_else(feature1 >= 0.3 & feature1 < 0.5,
                         sin(pi* feature1) + rnorm(n(), 0, 0.5),
                 if_else(feature1 >= 0.5 & feature1 < 0.8,
                         -(sin(pi* feature1) + rnorm(n(), 0, 0.1)) + 1,
                 if_else(feature1 >= 0.8 & feature1 < 0.9,
                         0.5 + runif(n(), -0.5, 0.5),
                         0.5 + rnorm(n(), 0, 0.3))))))) %>%
  bind_rows(
    data.frame(feature1 = rep(1, 30) + rnorm(5, 0, 0.05),
           feature2 = rep(245, 30) + rnorm(5, 0, 1),
           prediction = rep(0, 30) + rnorm(5, 0, 0.05)) %>%
  mutate(feature1_stnd = (feature1 - mean(feature1)) / sd(feature1),
         feature2_stnd = (feature2 - mean(feature2)) / sd(feature2)))

@@

\begin{figure*}[!thp]
<<figure-01, out.width = '6.5in', fig.width = 7.95, fig.height = 3.75, warning = FALSE>>=

# Specify the figure size (for determining font size)
f1_ow = 6.5
f1_fw = 7.95
f1_fs = fs * (f1_fw / f1_ow)

# Plot of good explainer model
ggplot() +
  facet_grid(. ~ feature, switch = "x") +
  geom_point(
    data = lime_data_gathered,
    mapping = aes(
      x = feature_stnd_value,
      y = prediction,
      size = distance_value,
      alpha = distance_value,
      color = distance
    )
  ) +
  geom_abline(
    data = coefs_data %>% filter(case == "good"),
    mapping = aes(intercept = int, slope = slope),
    size = 1
  ) +
  scale_alpha(range = c(0.3, 1)) +
  scale_color_manual(values = c(NA, "grey30"), guide = "none") +
  geom_point(
    data = prediction_of_interest_gathered,
    mapping = aes(x = feature_value,
                  y = prediction,
                  fill = color),
    color = "black",
    size = 5,
    shape = 23,
    alpha = 0.75
  ) +
  scale_fill_manual(values = "#FAAA72") +
  geom_text(
    data = data.frame(
      feature_stnd_value = 1.2,
      prediction = 1.6,
      feature = c("Feature 1", "Feature 2"),
      slope = c(
        paste(
          "Slope:",
          coefs_data %>%
            filter(case == "good",
                   feature == "Feature 1") %>%
            pull(slope) %>%
            round(3)
        ),
        paste(
          "Slope:",
          coefs_data %>%
            filter(case == "good",
                   feature == "Feature 2") %>%
            pull(slope) %>%
            round(3)
        )
      )
    ),
    mapping = aes(x = feature_stnd_value,
                  y = prediction,
                  label = slope),
    family = "Times",
    size = f1_fs / 2.85
  ) +
  labs(
    x = "",
    y = "Black-Box Prediction",
    title = "Conceptual Depiction of a Faithful Local Explainer Model",
    subtitle = paste("Gower Distance Metric Exponent:", good_gower_power),
    fill = "",
    alpha = "Weight",
    size = "Weight"
  ) +
  theme_linedraw(base_family = "Times", base_size = f1_fs) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    strip.placement = "outside",
    strip.background = element_rect(color = "white", fill = "white"),
    strip.text.x = element_text(size = f1_fs, color = "black"),
    plot.title = element_text(size = f1_fs)
  ) +
  guides(
    fill = guide_legend(order = 1),
    size = guide_legend(order = 2),
    alpha = guide_legend(order = 2)
  )

@
\caption{A conceptual depiction of a faithful LIME explainer model in the immediate neighborhood of a prediction of interest. The predictions from a hypothetical black-box model are plotted against the standardized values of the two hypothetical predictor variables. The diamond shaped points represent the location of a prediction of interest. The size and opacity of the circular points indicate the weight assigned based on the distance to the prediction of interest computed using the inverse of the Gower distance metric raised to the power of \Sexpr{good_gower_power}. The black lines represent  a weighted ridge regression model used as an explainer model that reasonably captures the relationship between the black-box predictions and the features in a local region around the prediction of interest. That is, the explainer is faithful to the complex model and produces a reasonable explanation that Feature 1 plays a more important role in the prediction of interest than Feature 2 since the magnitude of the slope associated with Feature 1 is larger.}
\label{fig:figure-01}

@@

<<figure-02, out.width = '6.5in', fig.width = 7.95, fig.height = 3.75, warning = FALSE>>=

# Specify the figure size (for determining font size)
f2_ow = 6.5
f2_fw = 7.95
f2_fs = fs * (f2_fw / f2_ow)

# Plot of good explainer model
ggplot() +
  facet_grid(. ~ feature, switch = "x") +
  geom_point(
    data = lime_data_gathered,
    mapping = aes(
      x = feature_stnd_value,
      y = prediction,
      size = distance_value,
      alpha = distance_value,
      color = distance
    )
  ) +
  geom_abline(
    data = coefs_data %>% filter(case == "bad"),
    mapping = aes(intercept = int, slope = slope),
    size = 1
  ) +
  scale_alpha(range = c(0.3, 1)) +
  scale_color_manual(values = c("grey30", NA), guide = "none") +
  geom_point(
    data = prediction_of_interest_gathered,
    mapping = aes(x = feature_value,
                  y = prediction,
                  fill = color),
    color = "black",
    size = 5,
    shape = 23,
    alpha = 0.75
  ) +
  scale_fill_manual(values = "#FAAA72") +
  geom_text(
    data = data.frame(
      feature_stnd_value = 1.2,
      prediction = 1.6,
      feature = c("Feature 1", "Feature 2"),
      slope = c(
        paste(
          "Slope:",
          coefs_data %>%
            filter(case == "bad",
                   feature == "Feature 1") %>%
            pull(slope) %>%
            round(3)
        ),
        paste(
          "Slope:",
          coefs_data %>%
            filter(case == "bad",
                   feature == "Feature 2") %>%
            pull(slope) %>%
            round(3)
        )
      )
    ),
    mapping = aes(x = feature_stnd_value,
                  y = prediction,
                  label = slope),
    family = "Times",
    size = f2_fs / 3
  ) +
  labs(
    x = "",
    y = "Black-Box Prediction",
    title = "Conceptual Depiction of an Unfaithful Local Explainer Model",
    subtitle = paste("Gower Distance Metric Exponent:", bad_gower_power),
    fill = "",
    alpha = "Weight",
    size = "Weight"
  ) +
  theme_linedraw(base_family = "Times", base_size = f2_fs) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    strip.placement = "outside",
    strip.background = element_rect(color = "white", fill = "white"),
    strip.text.x = element_text(size = f2_fs, color = "black"),
    plot.title = element_text(size = f2_fs)
  ) +
  guides(
    fill = guide_legend(order = 1),
    size = guide_legend(order = 2),
    alpha = guide_legend(order = 2)
  )

@
\caption{A conceptual depiction of an unfaithful local explainer model in the immediate neighborhood of a prediction of interest. A ridge regression model is fit to the same hypothetical black-box model predictions and predictor variables as in \autoref{fig:figure-01}, but the weights are computed using a Gower exponent of 1. The explainer model is unfaithful to the complex model in any immediate neighborhood of the prediction of interest.}
\label{fig:figure-02}
\end{figure*}

\autoref{fig:figure-01} provides a visualization of this conceptual understanding of LIME. The two plots show the predictions from a hypothetical black-box model plotted against the two hypothetical model predictor variables. The diamond shaped points represent the location of the prediction of interest. A Gower distance metric \citep{gower:1971} is used to define locality such that the size of the points represent the inverse of the Gower distance raised to some value and indicate the proximity to the prediction of interest. In the example of \autoref{fig:figure-01}, an exponent of \Sexpr{good_gower_power} is used to emphasize a very local region around the prediction of interest. A ridge regression model weighted by the proximity values is used as the explainer model with the black-box predictions as the response variable and standardized versions of two features in the data as predictor variables. Standardized features allow direct comparisons of the model coefficients. The explainer model is depicted by the black lines in the figure.

The plot on the left of \autoref{fig:figure-01} shows the relationship between the black-box predictions and Feature 1. Here, the explainer model is plotted with Feature 2 fixed at the observed value of the prediction of interest. The explainer model captures the relationship in an  immediate neighborhood around the prediction of interest with a slope of \Sexpr{coefs_data %>% filter(case == "good", feature == "Feature 1") %>% pull(slope) %>% round(3)}. The plot on the right shows no global or local relationships between the black-box predictions and Feature 2. Here, the explainer model is plotted with Feature 1 set as the observed value of the prediction of interest, and it has an appropriately small slope of \Sexpr{coefs_data %>% filter(case == "good", feature == "Feature 2") %>% pull(slope) %>% round(3)}. The magnitude of the slope associated with Feature 1 is larger than the slope of Feature 2, which suggests that Feature 1 plays a more important role in the prediction made by the black-box model for the prediction of interest. This explanation agrees with a visual assessment of the relationships between the predictions and predictor variables.

@@

We consider an example where the choice of exponent for the weights is crucial to the quality of the explanation. The plots in Figure \ref{fig:figure-02} show the same data as \autoref{fig:figure-01}, but the Gower distance metric exponent is decreased to \Sexpr{bad_gower_power} (the default exponent in the \emph{lime} R package). This causes the observations that are further away from the prediction of interest to be given larger weights than before. In \autoref{fig:figure-01}, the explainer model captures the relationship between the black-box predictions in an immediate neighborhood of the prediction of interest. This cannot be said of the explainer model in \autoref{fig:figure-02}. In addition, the magnitude of the slope associated with Feature 2  (\Sexpr{coefs_data %>% filter(case == "bad", feature == "Feature 2") %>% pull(slope) %>% round(3) %>% abs()}) is larger than that of Feature 1 (\Sexpr{coefs_data %>% filter(case == "bad", feature == "Feature 1") %>% pull(slope) %>% round(3) %>% abs()}). Thus, this explainer model actually provides a misleading explanation that Feature 2 plays a more important role in the prediction of interest.

Several sources in the literature discuss the performance of LIME. One of the biggest difficulties with LIME is determining how to specify a local region \citep{laugel:2018} \citep{molnar:2019}. This is due to an unclear definition of a "local region" and how to apply LIME to achieve an appropriate local region as demonstrated by \autoref{fig:figure-02}. \citet{alvarezmelis:2018} raise a concern pertaining to the robustness of explanations from LIME and other model explainers: they find that even small changes in predictor variables can lead to very different LIME explanations. Additionally, \citet{ribeiro:2016} acknowledge that if a linear model is used as the explainer, LIME relies on a linear approximation of the explainer model to the complex model and state "if the underlying model is highly non-linear even in the locality of the prediction, there may not be a faithful explanation".

@@

\begin{figure*}[!thp]
\centering
<<figure-03, out.width = '6.5in', fig.width = 8, fig.height = 3.1>>=

# Specify the figure size (for determining font size)
f3_ow = 6.5
f3_fw = 8
f3_fs = fs * (f3_fw / f3_ow)
f3_ls = 0.5 * (f3_fw / f3_ow)

@@

# Plot the training data observed classes
sine_plot_obs <- ggplot(sine_data_train, aes(x = x1, y = x2, color = y)) +
  geom_point(alpha = 0.8) +
  geom_path(data = sinefun_data, aes(x = x, y = y),
            color = "black", 
            size = f3_ls) +
  scale_colour_manual(values = c(sinecolor_blue, sinecolor_red)) +
  theme_bw(base_family = "Times", base_size = f3_fs) +
  theme(aspect.ratio = 1, 
        plot.title = element_text(size = f3_fs)) +
  labs(x = TeX("$x_1$"),
       y = TeX("$x_2$"),
       color = TeX("y"),
       title = "Training Data")

# Plot the testing data rf predictions
sine_plot_pred <- ggplot() +
  geom_point(data = sine_poi %>%
               mutate(shape = "Prediction \nof Interest"),
             mapping = aes(x = x1,
                           y = x2,
                           shape = shape),
             size = 5,
             alpha = 0.8,
             color = "black") +
  geom_point(data = sine_data_test %>% filter(y != rfpred),
             mapping = aes(x = x1, y = x2),
             shape = 1,
             size = 3,
             color = "black") +
  geom_point(data = sine_data_test,
             mapping =  aes(x = x1,
                            y = x2,
                            color = rfprob_blue,
                            fill = rfprob_blue),
             shape = 21,
             alpha = 0.8) +
  geom_path(data = sinefun_data, aes(x = x, y = y),
            color = "black", 
            size = f3_ls) +
  scale_color_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5) +
  scale_fill_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5) +
  scale_shape_manual(values = 23) +
  theme_bw(base_family = "Times", base_size = f3_fs) +
  theme(aspect.ratio = 1,
        plot.title = element_text(size = f3_fs)) +
  labs(x = TeX("$x_1$"),
       y = TeX("$x_2$"),
       color = "Random \nForest \nProbability \nfor 'blue'",
       fill = "Random \nForest \nProbability \nfor 'blue'",
       title = "Testing Data",
       shape = "") +
  guides(shape = guide_legend(order = 1))

@@

@
\caption{Plots of $x_2$ versus $x_1$ from the training (left) and testing (right) sets of the \data \ introduced in Section \ref{diagnostics}. The true classification boundary is shown as the solid black line in both plots. The color of the training data  represents the value of the observed response variable ($y$). The color of the testing data  represents the random forest probability that an observation belongs to the category of blue. The \Sexpr{dim(sine_data_test %>% filter(y != rfpred))[1]} cases that are misclassified by the random forest are identified by black circles. The prediction of interest to explain is indicated by a diamond.}
\label{fig:figure-03}
\end{figure*}

@@

# Extract the explanations from the default lime application
# for the prediction of interest only
sine_poi_lime_default <- sine_lime_default %>%
  filter(case == sine_poi %>% pull(case)) %>%
  mutate(case = c("Prediction of Interest", "Prediction of Interest"))

@@

\autoref{fig:figure-03} shows scatterplots of $x_2$ versus $x_1$ from the training data (left) and the testing data (right). Both plots include the true classification boundary of the rotated sine function plotted as the solid black line. The training data are colored by the observed response variable ($y$), and the testing data are colored by random forest  prediction probabilities. The random forest  misclassifies \Sexpr{dim(sine_data_test %>% filter(y != rfpred))[1]} points, which are all located near  the classification boundary. These are identified by open circles in \autoref{fig:figure-03}.

@@

For the presentation of the explanation scatterplot, we focus on the misclassified point with $(x_1, x_2, x_3)$ coordinates of (\Sexpr{round(sine_poi$x1, 2)}, \Sexpr{round(sine_poi$x2, 2)}, \Sexpr{round(sine_poi$x3, 2)}) indicated by a diamond in \autoref{fig:figure-03}. Misclassified points are often of interest to explain since they may provide information about ways to improve the model. For the introduction of the other two  plots, we consider the LIME explanations for  all observations in the \data \ test data.

<<warning = FALSE>>=

# Obtain the simulated data associated with the poi
sine_poi_perms <- sine_poi_lime_default %>%
  slice(1) %>%
  select(perms_raw, perms_pred_complex, perms_numerified, weights) %>%
  mutate(perms_numerified =
           map(perms_numerified,
               .f = function(x) rename(x,
                                       "x1num" = "x1",
                                       "x2num" = "x2",
                                       "x3num" = "x3"))) %>%
  unnest(cols = c(perms_raw, perms_pred_complex, perms_numerified, weights)) %>%
  select(-red) %>%
  rename(rfpred = blue) %>%
  mutate(case = factor(1:n())) %>%
  select(case, everything())

# Determine the bin cuts for the default lime application
sine_lime_default_bin_cuts <- sine_lime_default %>%
  select(feature, feature_desc) %>%
  unique() %>%
  separate(feature_desc, c("other", "bin_cut"), sep = " <= ") %>%
  select(-other) %>%
  na.omit() %>%
  mutate(bin_cut = as.numeric(bin_cut)) %>%
  arrange(feature) %>%
  mutate(case = rep(1:3, 2)) %>%
  spread(feature, bin_cut)

@@

An explanation scatterplot corresponding to the LIME explanation depicted on the left in \autoref{fig:figure-04} is shown on the right hand side of \autoref{fig:figure-04}. By juxtaposing the random forest predictions and the explainer model boundaries, we are able to assess the faithfulness and localness of the explainer model.
